The \emph{FThresholds} package for \emph{Macaulay2}

### 2019-06-04 Fsing/FThresholdsPaper/FThresholds.tex
More work on The fpt function'' section of the paper

The core function in this package is the {\tt fpt} function.  Given a polynomial $f$ with coefficients in a finite field, the function {\tt fpt} attempts to find the exact value for the $F$-pure threshold of $f$ at the origin, and returns that value, if possible.  Otherwise, it returns lower and upper bounds for the $F$-pure threshold, as demonstrated below.

\medskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i1 : ZZ/5[x,y,z];

@@

o3 : List
\end{MyVerbatim}
}

             If the option {\tt UseSpecialAlgorithms} is set to {\tt true} (the default value), then {\tt fpt} first checks whether $f$ is diagonal polynomial, a binomial, or a form in two variables, respectively.
             If it is one of these, algorithms of Hern\'andez, or Hern\'andez and Teixeira, {\color{red}  References?} are executed to compute the $F$-pure threshold of $f$.

\medskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i4 : fpt( x^17 + y^20 + z^24 ) -- a diagonal polynomial

@@

i7 : fpt( x^2*y^6*(x + y)^9*(x + 3*y)^10 ) -- a binary form in two variables

@@

o7 : QQ
\end{MyVerbatim}
}

            The computation of the $F$-pure threshold of a binary form $f$ requires factoring $f$ into linear forms, and can sometimes hang when attempting that factorization. For this reason, when a factorization is already known, the user can pass to {\tt fpt} a list containing all the pairwise prime linear factors of $f$ and a list containing their respective multiplicities.

\medskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i8 : L = {x, y, x + y, x + 3*y};

@@

o10 = true
\end{MyVerbatim}
}

The behavior of the {\tt fpt} function is more involved when no special algorithm is available, or {\tt UseSpecialAlgorithms} is set to {\tt false}.  In the remainder of this section, we describe the behavior of the {\tt fpt} function in this case, as well as the roles of the options {\tt DepthOfSearch}, and {\tt Attempts}, both nonnegative integers.

            When no special algorithm is available or , then function {\tt fpt} computes {\tt nu}$(e,f)$, where $e$ is the value of the option {\tt DepthOfSearch}, which conservatively defaults to $1$.
            At this point, we know that the $F$-pure threshold of $f$ lies in the closed interval {\color{red} Elaborate here} [$\nu/(p^e-1),(\nu+1)/p^e$], and the subroutine {\tt guessFPT} is called to make some "educated guesses" in an attempt to find the $F$-pure threshold, or at least narrow down the above interval.
	    The number of "guesses" is controlled by the option {\tt Attempts}, which conservatively defaults to 3.
            If {\tt Attempts} is set to 0, {\tt guessFPT} is bypassed.
            If  {\tt Attempts} is set to at least 1, then a first check is run to verify whether the right-hand endpoint $(\nu+1)/p^e$ of the above interval is the $F$-pure threshold.

\begin{example}
\[             f = x^2*(x + y)^3*(x + 3*y^2)^5;
            fpt( f, Attempts => 0 ) -- a bad estimate
            fpt( f, Attempts => 0, DepthOfSearch => 3 ) -- a better estimate
            fpt( f, Attempts => 1, DepthOfSearch => 3 ) -- the right-hand endpoint (nu+1)/p^e is the fpt \] 
\end{example}

            If  {\tt Attempts} is set to at least 2 and the right-hand endpoint $(\nu+1)/p^e$ is not the $F$-pure threshold, a second check is run to verify whether the left-hand endpoint $\nu/(p^e-1)$ is the $F$-pure threshold.

\begin{example}
\[
	    f = x^6*y^4 + x^4*y^9 + (x^2 + y^3)^3;
            fpt( f, Attempts => 1, DepthOfSearch => 3 )
            fpt( f, Attempts => 2, DepthOfSearch => 3 ) -- the left-hand endpoint nu/(p^e-1) is the fpt
\]
\end{example}

            If neither endpoint is the $F$-pure threshold, and {\tt Attempts} is set to more than 2, additional checks are performed at numbers in the interval.
            A number in the interval with minimal denominator is selected, and {\tt compareFPT} is used to test that number.
            If that ``guess" is correct, its value is returned; otherwise, the information returned by {\tt compareFPT} is used to narrow down the interval, and this process is repeated as many times as specified by {\tt Attempts}.

\begin{example}
\[             f = x^3*y^11*(x + y)^8*(x^2 + y^3)^8;
            fpt( f, DepthOfSearch => 3, Attempts => 2 )
            fpt( f, DepthOfSearch => 3, Attempts => 3 ) -- an additional check sharpens the estimate
            fpt( f, DepthOfSearch => 3, Attempts => 4 ) -- and one more finds the exact answer
\]
\end{example}

@@

\begin{example}
\[             f = x^5*y^6*(x + y)^9*(x^2 + y^3)^4;
            fpt( f, DepthOfSearch => 3 )
            fpt( f, DepthOfSearch => 3, UseFSignature => true )
            numeric ooo
            numeric ooo -- UseFSignature sharpened the estimate a bit \]
\end{example}

@@

\begin{example}
\[             f = (x + y)^4*(x^2 + y^3)^6;
            fpt( f, Attempts => 2, DepthOfSearch => 3 )
            fpt( f, Attempts => 2, DepthOfSearch => 3, UseFSignature => true ) -- using FSignatures the answer improves a bit
	    fpt( f, Attempts => 2, DepthOfSearch => 3, UseFSignature => true, FRegularityCheck => true ) -- FRegularityCheck finds the answer \] 
\end{example}

@@

\begin{example}            
\[            f = x^7*y^5*(x + y)^5*(x^2 + y^3)^4;
	    timing numeric fpt( f, DepthOfSearch => 3, UseFSignature => true, FRegularityCheck => true )
            timing numeric fpt( f, Attempts => 5, DepthOfSearch => 3 ) -- a better answer in less time
            timing fpt( f, DepthOfSearch => 4 ) -- the exact answer in even less time
\]
\end{example}

@@

\begin{example}
 \[           f = x^7*y^5*(x + y)^5*(x^2 + y^3)^4;
            fpt( f, DepthOfSearch => 3, UseFSignature => true, Verbose => true )
 \]
  \end{example}

>>>>

The core function in this package is the {\tt fpt} function.  Throughout this section, let $f$ be a polynomial with coefficients in a finite field of characteristic $p$. When passed the polynomial $f$, the function {\tt fpt} attempts to find the exact value for the $F$-pure threshold of $f$ at the origin, and returns that value, if possible.  Otherwise, it returns lower and upper bounds for the $F$-pure threshold, as demonstrated below.

{\small
\setstretch{.67}
\begin{MyVerbatim}

i1 : ZZ/5[x,y,z];

@@

o3 : List

\end{MyVerbatim}
}

             If the option {\tt UseSpecialAlgorithms} is set to {\tt true} (the default value), then {\tt fpt} first checks whether $f$ is diagonal polynomial, a binomial, or a form in two variables, in that order.
             If it is one of these, algorithms of Hern\'andez, or Hern\'andez and Teixeira, {\color{red}  References?} are executed to compute the $F$-pure threshold of $f$.

{\small
\setstretch{.67}
\begin{MyVerbatim}

i4 : fpt( x^17 + y^20 + z^24 ) -- a diagonal polynomial

@@

i7 : fpt( x^2*y^6*(x + y)^9*(x + 3*y)^10 ) -- a form in two variables

@@

o7 : QQ

\end{MyVerbatim}
}

            The above noted algorithm for computing the $F$-pure threshold of a binary form $f$ requires factoring $f$ into linear forms, and {\tt fpt} can sometimes hang when attempting that factorization. For this reason, when a factorization is already known, the user can pass to {\tt fpt} a list containing all the pairwise prime linear factors of $f$, and a list containing their respective multiplicities.

{\small
\setstretch{.67}
\begin{MyVerbatim}

i8 : L = {x, y, x + y, x + 3*y};

@@

o10 = true

\end{MyVerbatim}
}

In the remainder of this section, we describe the {\tt fpt} function when no special algorithm is available, or {\tt UseSpecialAlgorithms} is set to {\tt false},  as well as the roles of the options {\tt DepthOfSearch}, and {\tt Attempts}, both nonnegative integers.

In this case, very roughly speaking, the {\tt fpt} function will either compute, and output, the exact value of the $F$-pure threshold of $f$, and otherwise recursively compute a finite sequence of upper and lower bounds for this value, and output the last of these, which will be the tightest among all computed.  The value of the option {\tt DepthOfSearch} determines the precision of the initial set of bounds, and the option {\tt Attempts}, in conjunction with some various subroutines, determines whether, and how, to produce new, tighter bounds from the previous ones.  %We note that a pair of bounds will be produced only if the exact value of the $F$-pure threshold remains undetermined.

In more detail, let $e$ denote the value of the option {\tt DepthofSearch}, which conservatively defaults to $1$.  The {\tt fpt} function first computes $\nu=\nu_f(p^e)$, which agrees with the output of {\tt nu(e,f)}.  It is well-known that the $F$-pure threshold of $f$ is greater than $\nu/p^e$ and at most $(\nu+1)/p^e$, and applying  {\color{red} \cite[Proposition 4.2]{}} to this lower bound tells us that the $F$-pure threshold of $f$ must be at least $\nu/(p^e-1)$.  In summary, we know that the $F$-pure threshold of $f$ must lie in the closed interval 
%
\begin{equation}
\label{estimating-interval: e}
\tag{$\dagger$}
\left[ \frac{\nu}{p^e-1}, \frac{\nu+1}{p^e} \right].
\end{equation}

With these estimates in hand, the subroutine {\tt guessFPT} is called to make some ``educated guesses" in an attempt to identify the $F$-pure threshold within this interval, or at least narrow down this interval to produce improved estimates.  The number of ``guesses" is controlled by the option {\tt Attempts}, which conservatively defaults to 3.  If {\tt Attempts} is set to 0, then {\tt guessFPT} is bypassed. If  {\tt Attempts} is set to at least 1, then a first check is run to verify whether the right-hand endpoint $(\nu+1)/p^e$ of the above interval \eqref{estimating-interval: e} the $F$-pure threshold.

 We illustrate this below.

{\small
\setstretch{.67}
\begin{MyVerbatim}

i11 : f = x^2*(x + y)^3*(x + 3*y^2)^5;

i12 : fpt( f, Attempts => 0 ) -- a bad estimate

          1
o12 = {0, -}
          5

o12 : List

i13 : fpt( f, Attempts => 0, DepthOfSearch => 3 ) -- a better estimate

        21   22
o13 = {---, ---}
       124  125

o13 : List

i14 : fpt( f, Attempts => 1, DepthOfSearch => 3 ) -- the right-hand endpoint (nu+1)/p^e is the fpt

       22
o14 = ---
      125

o14 : QQ

\end{MyVerbatim}

If  {\tt Attempts} is set to at least 2, and the right-hand endpoint $(\nu+1)/p^e$ of the interval \eqref{estimating-interval: e} is not the $F$-pure threshold, then a second check is run to verify whether the left-hand endpoint $\nu/(p^e-1)$ of this interval is the $F$-pure threshold.

{\small
\setstretch{.67}
\begin{MyVerbatim}

i15 : f = x^6*y^4 + x^4*y^9 + (x^2 + y^3)^3;

i16 : fpt( f, Attempts => 1, DepthOfSearch => 3 )

       17   7
o16 = {--, --}
       62  25

o16 : List

i17 : fpt( f, Attempts => 2, DepthOfSearch => 3 ) -- the left-hand endpoint nu/(p^e-1) is the fpt

      17
o17 = --
      62

o17 : QQ

\end{MyVerbatim}
}

            If neither endpoint is the $F$-pure threshold, and {\tt Attempts} is set to more than 2, additional checks are performed at certain numbers within the interval .  A number in the interval \eqref{estimating-interval: e} with minimal denominator is selected, and {\tt compareFPT} is used to test that number. If that ``guess" is correct, its value is returned; otherwise, the information returned by {\tt compareFPT} is used to narrow down the interval, and this process is repeated as many times as specified by {\tt Attempts}.

{\small
\setstretch{.67}
\begin{MyVerbatim}

i18 : f = x^3*y^11*(x + y)^8*(x^2 + y^3)^8;

i19 : fpt( f, DepthOfSearch => 3, Attempts => 2 )

        3   7
o19 = {--, ---}
       62  125

o19 : List

i20 : fpt( f, DepthOfSearch => 3, Attempts => 3 ) -- an additional check sharpens the estimate

        3   1
o20 = {--, --}
       62  18

o20 : List

i21 : fpt( f, DepthOfSearch => 3, Attempts => 4 ) -- and one more finds the exact answer

       1
o21 = --
      19

o21 : QQ

\end{MyVerbatim}
}

@@

{\small
\setstretch{.67}
\begin{MyVerbatim}

i22 : f = x^5*y^6*(x + y)^9*(x^2 + y^3)^4;

i23 : fpt( f, DepthOfSearch => 3 )

        2   1
o23 = {--, --}
       31  14

o23 : List

i24 : fpt( f, DepthOfSearch => 3, UseFSignature => true )

        181   1
o24 = {----, --}
       2750  14

o24 : List

i25 : numeric o23

o25 = {.0645161, .0714286}

o25 : List

i26 : numeric o24 -- UseFSignature sharpened the estimate a bit

o26 = {.0658182, .0714286}

o26 : List

\end{MyVerbatim}
}

@@

{\small
\setstretch{.67}
\begin{MyVerbatim}

i27 : f = (x + y)^4*(x^2 + y^3)^6;

i28 : fpt( f, Attempts => 2, DepthOfSearch => 3 )

        3   13
o28 = {--, ---}
       31  125

o28 : List

i29 : fpt( f, Attempts => 2, DepthOfSearch => 3, UseFSignature => true ) -- using FSignatures the answer improves a bit

        1   13
o29 = {--, ---}
       10  125

o29 : List

i30 : fpt( f, Attempts => 2, DepthOfSearch => 3, UseFSignature => true, FRegularityCheck => true ) -- FRegularityCheck finds the answer

       1
o30 = --
      10

o30 : QQ

\end{MyVerbatim}
}

@@

{\small
\setstretch{.67}
\begin{MyVerbatim}

i31: f = x^7*y^5*(x + y)^5*(x^2 + y^3)^4;

i32 : timing numeric fpt( f, DepthOfSearch => 3, UseFSignature => true, FRegularityCheck => true )

o32 = {.0733061, .0769231}
      -- 2.68899 seconds

o32 : Time

i33 : timing numeric fpt( f, Attempts => 5, DepthOfSearch => 3 ) -- a better answer in less time

o33 = {.075, .0769231}
      -- .893389 seconds

o33 : Time

i34 : timing fpt( f, DepthOfSearch => 4 ) -- the exact answer in even less time

       48
o34 = ---
      625
      -- .361882 seconds

o34 : Time

\end{MyVerbatim}
}

@@

{\small
\setstretch{.67}
\begin{MyVerbatim}

i35 : f = x^7*y^5*(x + y)^5*(x^2 + y^3)^4;

i36 :  fpt( f, DepthOfSearch => 3, UseFSignature => true, Verbose => true )

Starting fpt ...

fpt is not 1 ...

Verifying if special algorithms apply...

Special fpt algorithms were not used ...

nu has been computed: nu = nu(3,f) = 9 ...

fpt lies in the interval [ nu/(p^e-1), (nu+1)/p^e ] = [ 9/124, 2/25 ] ...

Starting guessFPT ...

The right-hand endpoint is not the fpt ...

The left-hand endpoint is not the fpt ...

guessFPT narrowed the interval down to ( 9/124, 1/13 ) ...

Beginning F-signature computation ...

First F-signature computed: s(f,(nu-1)/p^e) = 456/15625 ...

Second F-signature computed: s(f,nu/p^e) = 64/15625 ...

Computed F-signature secant line intercept: 449/6125 ...

F-signature intercept is an improved lower bound ...

fpt failed to find the exact answer; try increasing the value of DepthOfSearch or Attempts.

fpt lies in the interval [ 449/6125, 1/13 ).

        449   1
o36 = {----, --}
       6125  13

o36 : List

\end{MyVerbatim}
}
